---
title: "EDA: MTBS Wildfire Perimeters — Western US"
subtitle: "Exploratory data analysis, pre-biomass-extraction phase"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 7
    theme: flatly
execute:
  warning: false
  message: false
  cache: false
---

```{r}
# =============================================================================
# SCRIPT OUTLINE
# 1. Setup — packages, paths, study parameters
# 2. Load MTBS data + sanity checks
# 3. Filter to study region (11 Western states) and period (2000–2023)
# 4. Plot 1 — Fire locations map (spatial distribution, colored by assessment type)
# 5. Plot 2 — Fires per year (temporal trend)
# 6. Plot 3 — Fire size distribution
# 7. Plot 4 — Severity threshold distributions (dNBR thresholds + dNBR offset)
# 8. Step 3 — GEE biomass extraction (run Python script if CSV missing)
# 9. Load and validate biomass_timeseries.csv
# 10. Plot 5 — Annual mean NBR over time (1984–2023)
# 11. Plot 6 — Biomass before and after fire events (event-study)
#
# NOTE ON SEVERITY FIELDS:
# The MTBS perimeter shapefile contains Low_T / Mod_T / High_T as dNBR THRESHOLD
# VALUES (classification breakpoints), NOT area percentages. Per-class area
# breakdowns (% low / moderate / high) require individual fire severity rasters
# which are not included in this download. See MTBS metadata for field definitions.
# =============================================================================
```

## 1. Setup

```{r setup}
# Install any missing packages before loading
.install_if_missing <- function(pkgs) {
  missing <- pkgs[!pkgs %in% rownames(installed.packages())]
  if (length(missing) > 0) {
    message("Installing missing packages: ", paste(missing, collapse = ", "))
    install.packages(missing, repos = "https://cloud.r-project.org", quiet = TRUE)
  }
}
.install_if_missing(c("tigris", "ggspatial", "glue", "patchwork", "scales", "png"))

library(tidyverse)         # data wrangling + ggplot2
library(sf)                # spatial vector data
library(tigris)            # US Census state boundaries
library(ggspatial)         # scale bar + north arrow
library(rnaturalearth)     # fallback state boundaries (already installed)
library(here)              # project-relative paths
library(scales)            # axis formatting (comma, log)
library(lubridate)         # date parsing
library(glue)              # string interpolation
library(patchwork)         # combine plots (Plots 5+6)
library(png)               # resolution check for saved figures

options(tigris_use_cache = TRUE)  # cache Census downloads locally

# Use GEOS (planar) instead of s2 (spherical) for spatial operations.
# MTBS perimeters contain self-touching edges that s2 rejects but GEOS handles
# correctly. For coterminous US data the planar approximation is fine.
sf_use_s2(FALSE)

# Anchor here() to the project root (one level up from analysis/)
here::i_am("analysis/01_exploration.qmd")

# --- Study parameters (all hardcoded values in one place) -------------------
WESTERN_STATES <- c("AZ", "CA", "CO", "ID", "MT", "NV", "NM", "OR", "UT", "WA", "WY")
YEAR_MIN       <- 2000
YEAR_MAX       <- 2023
MIN_FIRE_ACRES <- 1000   # MTBS default minimum (fires <1000 ac are not mapped)
MTBS_PATH      <- here("data/raw/mtbs/mtbs_perimeter_data/mtbs_perims_DD.shp")

# --- Assessment type color palette (colorblind-friendly) --------------------
# Fires are colored by mapping assessment type (Initial vs Extended), since the
# perimeter shapefile has no per-class area data — only dNBR threshold values.
ASMT_COLORS <- c(
  "Initial"          = "#4393c3",
  "Initial (SS)"     = "#92c5de",
  "Extended"         = "#d6604d",
  "Extended (SS)"    = "#f4a582",
  "Emergency"        = "#878787",
  "Emergency (SS)"   = "#bababa"
)

# --- Output directory --------------------------------------------------------
dir.create(here("figures"), showWarnings = FALSE)
```

---

## 2. Load and Inspect MTBS Data

```{r load-mtbs}
# Load the raw shapefile
mtbs_raw <- st_read(MTBS_PATH, quiet = TRUE)

# --- Sanity checks on raw data ----------------------------------------------
# IMPORTANT: low_t / mod_t / high_t are dNBR THRESHOLD VALUES (index units),
# NOT area percentages. They define the dNBR breakpoints between severity classes
# and vary fire-by-fire based on analyst calibration. Do not compare them as if
# they were proportions of burned area.
stopifnot(
  "Missing expected fields" =
    all(c("event_id", "ig_date", "incid_type", "burnbndac",
          "low_t", "mod_t", "high_t", "dnbr_offst", "asmnt_type") %in% names(mtbs_raw))
)

# Check geometry validity and fix if needed
n_invalid <- sum(!st_is_valid(mtbs_raw))
if (n_invalid > 0) {
  message(glue::glue("Fixing {n_invalid} invalid geometries with st_make_valid()"))
  mtbs_raw <- st_make_valid(mtbs_raw)
}

cat("Records loaded:   ", nrow(mtbs_raw), "\n")
cat("Incident types:   ", paste(sort(unique(mtbs_raw$incid_type)), collapse = ", "), "\n")
cat("CRS:              ", st_crs(mtbs_raw)$input, "\n")
```

```{r parse-fields}
# Parse year from ig_date ("YYYY-MM-DD" character string)
# Flag fires with valid (non-sentinel) severity thresholds.
# Sentinel values -9999 and 9999 mean "No Data" per the MTBS data dictionary.
mtbs_parsed <- mtbs_raw |>
  mutate(
    year = as.integer(substr(ig_date, 1, 4)),
    # TRUE when all three threshold fields are populated with real dNBR values
    thresholds_valid = !(low_t %in% c(-9999, 9999) |
                         mod_t %in% c(-9999, 9999) |
                         high_t %in% c(-9999, 9999))
  )

cat("Year range in raw data:", min(mtbs_parsed$year, na.rm = TRUE),
    "–", max(mtbs_parsed$year, na.rm = TRUE), "\n")
cat("Fires with valid thresholds:", sum(mtbs_parsed$thresholds_valid),
    "/", nrow(mtbs_parsed), "\n")
```

---

## 3. Filter to Study Region and Period

```{r get-state-boundaries}
# Download Western US state boundaries (cb = TRUE for simplified cartographic)
western_states_sf <- states(cb = TRUE, progress_bar = FALSE) |>
  filter(STUSPS %in% WESTERN_STATES) |>
  st_transform(st_crs(mtbs_parsed))   # match CRS to MTBS

cat("State boundaries loaded for:", paste(sort(western_states_sf$STUSPS), collapse = ", "), "\n")
stopifnot("Not all 11 states returned" = nrow(western_states_sf) == 11)
```

```{r filter-region-period}
# Step 1: Wildfire only (exclude Prescribed Fire)
n_before <- nrow(mtbs_parsed)
mtbs_wf <- mtbs_parsed |> filter(incid_type == "Wildfire")
cat("After Wildfire filter:        ", nrow(mtbs_wf), "(removed", n_before - nrow(mtbs_wf), "prescribed fires)\n")

# Step 2: Year filter (2000–2023, matching biomass data availability)
mtbs_wf <- mtbs_wf |> filter(year >= YEAR_MIN & year <= YEAR_MAX)
cat("After year filter (2000–23):  ", nrow(mtbs_wf), "\n")

# Step 3: Spatial intersect with 11 Western states
# Use st_join to attach state abbreviation, then drop fires outside study region
mtbs_west <- st_join(mtbs_wf, western_states_sf["STUSPS"], join = st_intersects, left = FALSE)
cat("After Western US spatial filter:", nrow(mtbs_west), "\n")

# Sanity check: confirm all 11 states are represented
present_states <- sort(unique(mtbs_west$STUSPS))
missing_states  <- setdiff(WESTERN_STATES, present_states)
if (length(missing_states) > 0) {
  warning(paste("States with no fires:", paste(missing_states, collapse = ", ")))
} else {
  cat("All 11 Western states have ≥1 fire. ✓\n")
}

# Final fire count for inline reference
n_fires_total <- nrow(mtbs_west)
cat("\nFinal dataset:", n_fires_total, "wildfires, Western US, 2000–2023\n")
```

```{r severity-sanity}
# ---- Sanity checks on dNBR threshold fields --------------------------------
# These checks guard against the most common misuse: treating threshold values
# (dNBR index units, typically 50–600) as if they were percentages (0–100).

valid_thresh <- mtbs_west |>
  st_drop_geometry() |>
  filter(thresholds_valid)

# Check 1: Threshold orientation.
# MTBS fires can be mapped on either axis of the dNBR spectrum:
#   "Burned" fires:             low_t < mod_t < high_t (thresholds increase toward high severity)
#   "Increased Greenness" fires: low_t > mod_t > high_t (thresholds decrease, negative dNBR)
# Both orderings are valid. What is NOT valid is a MIXED ordering (non-monotone),
# which would indicate corrupted threshold values.
valid_thresh <- valid_thresh |>
  mutate(
    thresh_orient = case_when(
      low_t < mod_t & mod_t < high_t ~ "burned",       # typical positive-dNBR fire
      low_t > mod_t & mod_t > high_t ~ "greenness",    # increased-greenness fire
      TRUE                            ~ "mixed"         # non-monotone — data problem
    )
  )

orient_tbl <- valid_thresh |> count(thresh_orient)
cat("Threshold orientation breakdown:\n")
print(orient_tbl)

n_mixed <- sum(valid_thresh$thresh_orient == "mixed")
if (n_mixed > 0) {
  warning(sprintf("%d fires have non-monotone thresholds (low_t, mod_t, high_t not ordered) — inspect for data issues.", n_mixed))
} else {
  cat("All valid fires have monotone thresholds (burned or increased-greenness). ✓\n")
}

# For severity distribution plots, restrict to "burned" orientation fires
# (positive dNBR axis), which is the relevant class for this biomass study.
sev_burned <- valid_thresh |> filter(thresh_orient == "burned")

# Check 2: Values must NOT all fall in 0–100 — that would mean we have percentages
# loaded into threshold fields (e.g. wrong join or wrong shapefile).
stopifnot(
  "high_t values all in 0–100 range — are these percentages, not dNBR thresholds?" =
    !(all(valid_thresh$high_t >= 0 & valid_thresh$high_t <= 100)),
  "low_t values all in 0–100 range — are these percentages, not dNBR thresholds?" =
    !(all(valid_thresh$low_t >= 0 & valid_thresh$low_t <= 100))
)
cat("Threshold scale check passed (values outside 0–100 percentage range). ✓\n")

# Check 3: Non-zero variance — if all fires had identical thresholds it would
# suggest a bad join, a constant-filled field, or the wrong column entirely.
stopifnot(
  "high_t has zero variance — all fires have same threshold (wrong field?)" =
    sd(valid_thresh$high_t, na.rm = TRUE) > 0
)
cat("Threshold variance check passed. ✓\n")

# Report no-data counts and value ranges
n_nodata <- sum(!mtbs_west$thresholds_valid, na.rm = TRUE)
cat("\nFires with no-data thresholds (-9999 or 9999):", n_nodata,
    sprintf("(%.1f%%)\n", 100 * n_nodata / nrow(mtbs_west)))

cat("\ndNBR threshold value ranges (valid fires only):\n")
cat("  High_T:", range(valid_thresh$high_t), "\n")
cat("  Mod_T: ", range(valid_thresh$mod_t),  "\n")
cat("  Low_T: ", range(valid_thresh$low_t),  "\n")

# Report dNBR offset range (proxy for background spectral change)
valid_offst <- mtbs_west |> st_drop_geometry() |>
  filter(!dnbr_offst %in% c(-9999, 9999))
cat("\ndNBR offset range (valid fires):", range(valid_offst$dnbr_offst), "\n")
```

::: callout-note
**Total wildfires in study region:** `r n_fires_total` fires (Western US, 2000–2023, wildfires ≥1,000 acres)

**Note on severity data:** The MTBS perimeter shapefile contains `Low_T`, `Mod_T`, and `High_T` as dNBR index threshold values — not area percentages. Per-class burn area distributions (% low / moderate / high) require individual fire severity rasters and are not available in this download.
:::

---

## Plot 1 — Fire Locations Map

Spatial distribution of MTBS wildfire perimeters across the 11 Western states, colored by mapping assessment type (Initial vs Extended). Initial assessments use imagery shortly after containment; Extended assessments use a later peak-greenness post-fire image for more accurate severity calibration. This map confirms spatial coverage and that the state filter worked correctly.

```{r plot1-map, fig.height=9}
# Use Albers Equal Area projection for the map display — avoids distortion
# across the latitudinal spread of the Western US
crs_albers <- 5070  # EPSG:5070 (Conus Albers Equal Area)

western_states_alb <- st_transform(western_states_sf, crs_albers)
mtbs_west_alb      <- st_transform(mtbs_west, crs_albers)

# Normalize assessment type labels to the palette keys (some may have trailing
# whitespace or alternate capitalizations in the shapefile)
mtbs_west_alb <- mtbs_west_alb |>
  mutate(asmnt_type = trimws(asmnt_type))

# Use only the assessment types actually present; fall back to grey for unknowns
present_asmnt <- sort(unique(mtbs_west_alb$asmnt_type))
cat("Assessment types present:", paste(present_asmnt, collapse = ", "), "\n")

p1 <- ggplot() +
  # State boundaries as base layer
  geom_sf(data = western_states_alb,
          fill = "grey96", color = "grey50", linewidth = 0.4) +
  # Fire perimeters, semi-transparent so overlapping fires are visible
  geom_sf(data = mtbs_west_alb,
          aes(fill = asmnt_type),
          color = NA, alpha = 0.65) +
  scale_fill_manual(
    name   = "Assessment type",
    values = ASMT_COLORS,
    na.value = "#cccccc"
  ) +
  # Scale bar (bottom-left) and north arrow (top-right)
  annotation_scale(location = "bl", width_hint = 0.25, text_cex = 0.85) +
  annotation_north_arrow(
    location = "tr", which_north = "true",
    style    = north_arrow_minimal(text_size = 10)
  ) +
  labs(
    title    = "MTBS Wildfire Perimeters — Western United States (2000–2023)",
    subtitle = glue::glue("{scales::comma(n_fires_total)} wildfires ≥ 1,000 acres"),
    caption  = "Source: USGS Monitoring Trends in Burn Severity (MTBS) Program"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(color = "grey40", margin = margin(b = 6)),
    plot.caption    = element_text(color = "grey50", size = 9),
    legend.position = "bottom",
    legend.title    = element_text(size = 11, face = "bold"),
    legend.text     = element_text(size = 10),
    plot.margin     = margin(10, 10, 10, 10)
  ) +
  guides(fill = guide_legend(nrow = 2, override.aes = list(alpha = 1)))

print(p1)

# --- QA: confirm legend is visible and plot non-empty -----------
stopifnot("Plot 1 has no data" = nrow(mtbs_west_alb) > 0)

ggsave(here("figures/fire_locations_map.png"),
       plot = p1, width = 10, height = 9, dpi = 300, bg = "white")

# Confirm output resolution
img_dims <- dim(png::readPNG(here("figures/fire_locations_map.png")))
cat("Plot 1 saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

## Plot 2 — Fires Per Year (Temporal Trend)

Annual wildfire count across the Western US, 2000–2023, with a linear trend line. This reveals whether fire frequency is increasing over time — a key consideration for the staggered DiD design (more treatment cohorts = better identification).

```{r plot2-fires-per-year}
fires_per_year <- mtbs_west |>
  st_drop_geometry() |>
  count(year, name = "n_fires")

# Sanity check: should have one row per year across the full range
expected_years <- YEAR_MIN:YEAR_MAX
missing_years  <- setdiff(expected_years, fires_per_year$year)
if (length(missing_years) > 0) {
  message("Years with zero fires (will add as 0): ", paste(missing_years, collapse = ", "))
  fires_per_year <- fires_per_year |>
    complete(year = expected_years, fill = list(n_fires = 0))
}

p2 <- ggplot(fires_per_year, aes(x = year, y = n_fires)) +
  geom_col(fill = "#fc8d59", color = "white", linewidth = 0.3, alpha = 0.85) +
  # Linear trend line to visualize direction of change
  geom_smooth(method = "lm", se = TRUE,
              color = "#d73027", fill = "#d73027", alpha = 0.15,
              linewidth = 1) +
  scale_x_continuous(
    breaks = seq(YEAR_MIN, YEAR_MAX, by = 2),
    expand = c(0.01, 0)
  ) +
  scale_y_continuous(
    labels = comma,
    expand = c(0, 0),
    limits = c(0, max(fires_per_year$n_fires) * 1.15)
  ) +
  labs(
    title   = "Annual Wildfire Count — Western United States",
    x       = "Year",
    y       = "Number of wildfires",
    caption = "Source: USGS MTBS. Wildfires ≥ 1,000 acres only. Red band = 95% CI around linear trend."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.caption    = element_text(color = "grey50", size = 9),
    panel.grid.minor = element_blank(),
    axis.title      = element_text(size = 11),
    axis.text       = element_text(size = 10)
  )

print(p2)

# --- QA ---
stopifnot("Y-axis must not be all zeros" = max(fires_per_year$n_fires) > 0)
stopifnot("X-axis must span 2000-2023"   = range(fires_per_year$year) == c(YEAR_MIN, YEAR_MAX))

ggsave(here("figures/fires_per_year.png"),
       plot = p2, width = 10, height = 6, dpi = 300, bg = "white")

img_dims <- dim(png::readPNG(here("figures/fires_per_year.png")))
cat("Plot 2 saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

## Plot 3 — Fire Size Distribution

Histogram of fire sizes (acres) on a log₁₀ scale. The log scale is necessary because fire sizes span several orders of magnitude. This informs the minimum fire size threshold decision — at what size does the biomass signal become detectable?

```{r plot3-fire-size}
# Summary stats to annotate the plot
size_stats <- mtbs_west |>
  st_drop_geometry() |>
  summarise(
    median_ac = median(burnbndac, na.rm = TRUE),
    mean_ac   = mean(burnbndac, na.rm = TRUE),
    p95_ac    = quantile(burnbndac, 0.95, na.rm = TRUE)
  )

cat("Fire size summary:\n")
cat("  Median:", scales::comma(size_stats$median_ac), "acres\n")
cat("  Mean:  ", scales::comma(round(size_stats$mean_ac)), "acres\n")
cat("  95th %:", scales::comma(size_stats$p95_ac), "acres\n")

# Sanity check: no negative or zero acre values
n_bad <- sum(mtbs_west$burnbndac <= 0, na.rm = TRUE)
if (n_bad > 0) warning(paste(n_bad, "fires have burnbndac <= 0"))

p3 <- ggplot(mtbs_west |> st_drop_geometry(),
             aes(x = burnbndac)) +
  geom_histogram(bins = 60, fill = "#fc8d59", color = "white",
                 linewidth = 0.2, alpha = 0.9) +
  # Mark median with a vertical line
  geom_vline(xintercept = size_stats$median_ac,
             color = "#d73027", linewidth = 1, linetype = "dashed") +
  annotate("text",
           x = size_stats$median_ac * 1.5,
           y = Inf, vjust = 1.5, hjust = 0, size = 3.5, color = "#d73027",
           label = paste0("Median: ", scales::comma(size_stats$median_ac), " ac")) +
  scale_x_log10(
    labels = label_comma(),
    breaks = c(1000, 5000, 10000, 50000, 100000, 500000)
  ) +
  scale_y_continuous(labels = comma, expand = c(0, 0)) +
  labs(
    title   = "Distribution of Wildfire Sizes — Western United States (2000–2023)",
    x       = "Fire size (acres, log₁₀ scale)",
    y       = "Number of fires",
    caption = "Source: USGS MTBS. Dashed line = median fire size."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.caption     = element_text(color = "grey50", size = 9),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(size = 11),
    axis.text        = element_text(size = 10)
  )

print(p3)

ggsave(here("figures/fire_size_distribution.png"),
       plot = p3, width = 10, height = 6, dpi = 300, bg = "white")

img_dims <- dim(png::readPNG(here("figures/fire_size_distribution.png")))
cat("Plot 3 saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

## Plot 4 — Burn Severity: dNBR and RdNBR Distributions

```{r plot4-severity}
# =============================================================================
# BURN SEVERITY METRICS — dNBR vs. RdNBR
#
# dNBR  (differenced Normalized Burn Ratio):
#   dNBR = (preFireNBR − postFireNBR) × 1000
#   Captures spectral change caused by fire. Positive values = vegetation
#   loss (burning). Stored in MTBS/BSP as integer index units (×1000 scale),
#   so values typically range from −500 (increased greenness) to 1300+ (high
#   severity). A useful absolute measure of spectral change per fire pixel.
#
#   LIMITATION: pre-fire biomass confounds cross-fire comparison. A dNBR of
#   300 in dense old-growth forest represents less relative canopy destruction
#   than 300 in sparse shrubland, because old-growth had more NBR to lose.
#
# RdNBR (Relativized dNBR; Miller & Thode 2007):
#   RdNBR = dNBR / sqrt(|preFireNBR / 1000|)
#   Divides by the square root of pre-fire NBR (rescaled to the 0–1 range),
#   normalizing for vegetation density before the fire. This makes severity
#   more comparable across fires in different biomes and vegetation types —
#   the preferred continuous treatment variable for a cross-fire DiD study.
#
# Data: USGS Burn Severity Program (BSP) CBI field plots — geospatially
# validated measurement points sampled within fire perimeters by field crews.
# BSP stores raw dNBR and pre-fire NBR per plot, enabling RdNBR computation.
# These are the only per-point raw dNBR observations available in this
# download without fetching individual MTBS fire severity rasters.
# =============================================================================

BSP_PATH <- here("data/raw/mtbs/burn_severity_perimeter_data/bsp_perims_DD.shp")

bsp_raw <- st_read(BSP_PATH, quiet = TRUE)
names(bsp_raw) <- tolower(names(bsp_raw))

# Keep only CBI field plots with valid dNBR and pre-fire NBR.
# Exclude sentinel 9999 (no data) and |preNBR| ≤ 10 (avoids RdNBR blow-up
# from near-zero pre-fire NBR in bare/sparse areas).
cbi_raw <- bsp_raw |>
  filter(
    map_prog   == "CBI",
    !is.na(dnbr_val),  !is.na(prenbr_val),
    dnbr_val   != 9999, prenbr_val != 9999,
    abs(prenbr_val) > 10
  ) |>
  mutate(
    # RdNBR: normalize by sqrt(|preNBR|) on the 0–1 NBR scale
    rdnbr = dnbr_val / sqrt(abs(prenbr_val / 1000))
  )

# Spatial filter to Western US (BSP is NAD83, matches western_states_sf CRS)
cbi_raw  <- st_transform(cbi_raw, st_crs(western_states_sf))
cbi_west <- st_filter(cbi_raw, western_states_sf)

cat("CBI field plots (Western US, valid dNBR + preNBR):", nrow(cbi_west), "\n")
cat("dNBR  range:", range(cbi_west$dnbr_val),            "\n")
cat("RdNBR range:", round(range(cbi_west$rdnbr, na.rm = TRUE), 1), "\n")

# --- Sanity checks -----------------------------------------------------------
stopifnot("No CBI plots in Western US"         = nrow(cbi_west) > 0)
stopifnot("All dNBR values identical"          = sd(cbi_west$dnbr_val) > 0)
n_rdnbr_na <- sum(is.na(cbi_west$rdnbr) | !is.finite(cbi_west$rdnbr))
if (n_rdnbr_na > 0)
  message(n_rdnbr_na, " RdNBR values are NA/Inf (pre-fire NBR near zero)")

# For mapping: extract centroids from mixed-geometry BSP features
cbi_pts_alb <- cbi_west |>
  st_centroid() |>
  st_transform(crs_albers)   # crs_albers defined in Plot 1 chunk

# Color scale limits: clip to 2nd–98th percentile to prevent outliers from
# collapsing the colour range
dnbr_lims <- quantile(cbi_west$dnbr_val, c(0.02, 0.98), na.rm = TRUE)

# --- 4a: Map — CBI field plots colored by dNBR -------------------------------
p4a <- ggplot() +
  geom_sf(data = western_states_alb,
          fill = "grey96", color = "grey50", linewidth = 0.4) +
  geom_sf(data = mtbs_west_alb,
          fill = "grey80", color = NA, alpha = 0.45) +
  geom_sf(data = cbi_pts_alb,
          aes(color = dnbr_val), size = 1.1, alpha = 0.8) +
  scale_color_viridis_c(
    name      = "dNBR\n(index units)",
    option    = "inferno",
    direction = -1,
    limits    = dnbr_lims,
    oob       = scales::squish,
    labels    = comma
  ) +
  annotation_scale(location = "bl", width_hint = 0.25, text_cex = 0.85) +
  annotation_north_arrow(
    location = "tr", which_north = "true",
    style    = north_arrow_minimal(text_size = 10)
  ) +
  labs(
    title    = "BSP CBI Field Plot Locations — Colored by Raw dNBR",
    subtitle = glue("{scales::comma(nrow(cbi_west))} validated field plots, Western US (1994–2017)"),
    caption  = paste0(
      "Grey polygons = MTBS fire perimeters. ",
      "Source: USGS Burn Severity Program (BSP) CBI field plots. ",
      "Color clipped to 2nd–98th percentile."
    )
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey40", margin = margin(b = 6)),
    plot.caption  = element_text(color = "grey50", size = 9),
    legend.position = "bottom",
    legend.title  = element_text(size = 11, face = "bold"),
    legend.text   = element_text(size = 10),
    plot.margin   = margin(10, 10, 10, 10)
  ) +
  guides(color = guide_colorbar(barwidth = 12, barheight = 0.8))

# --- 4b: Histogram — dNBR distribution --------------------------------------
dnbr_med <- median(cbi_west$dnbr_val, na.rm = TRUE)

p4b <- ggplot(cbi_west |> st_drop_geometry(), aes(x = dnbr_val)) +
  geom_histogram(bins = 60, fill = "#d73027", color = "white",
                 linewidth = 0.2, alpha = 0.85) +
  geom_vline(xintercept = 0,
             color = "black", linewidth = 0.6, linetype = "dotted") +
  geom_vline(xintercept = dnbr_med,
             color = "#4d4d4d", linewidth = 1, linetype = "dashed") +
  annotate("text",
           x = dnbr_med + 15, y = Inf, vjust = 1.5, hjust = 0,
           size = 3.5, color = "#4d4d4d",
           label = paste0("Median: ", round(dnbr_med))) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma, expand = c(0, 0)) +
  labs(
    title   = "Distribution of Raw dNBR at BSP CBI Field Plots — Western US",
    x       = "dNBR (index units;  positive = burning,  negative = increased greenness)",
    y       = "Number of field plots",
    caption = paste0(
      "Dotted line = 0 (no change). Dashed line = median. ",
      "n = ", scales::comma(nrow(cbi_west)), " CBI field plots."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.caption     = element_text(color = "grey50", size = 9),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(size = 11),
    axis.text        = element_text(size = 10)
  )

# --- 4c: Histogram — RdNBR distribution -------------------------------------
rdnbr_df  <- cbi_west |> st_drop_geometry() |>
  filter(!is.na(rdnbr), is.finite(rdnbr))
rdnbr_med <- median(rdnbr_df$rdnbr, na.rm = TRUE)

p4c <- ggplot(rdnbr_df, aes(x = rdnbr)) +
  geom_histogram(bins = 60, fill = "#4393c3", color = "white",
                 linewidth = 0.2, alpha = 0.85) +
  geom_vline(xintercept = 0,
             color = "black", linewidth = 0.6, linetype = "dotted") +
  geom_vline(xintercept = rdnbr_med,
             color = "#4d4d4d", linewidth = 1, linetype = "dashed") +
  annotate("text",
           x = rdnbr_med + 10, y = Inf, vjust = 1.5, hjust = 0,
           size = 3.5, color = "#4d4d4d",
           label = paste0("Median: ", round(rdnbr_med))) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma, expand = c(0, 0)) +
  labs(
    title   = "Distribution of RdNBR at BSP CBI Field Plots — Western US",
    x       = "RdNBR (relativized dNBR; normalized for pre-fire vegetation density)",
    y       = "Number of field plots",
    caption = paste0(
      "RdNBR = dNBR / sqrt(|preNBR / 1000|). ",
      "Dotted line = 0. Dashed line = median. ",
      "n = ", scales::comma(nrow(rdnbr_df)), " CBI field plots. ",
      "Ref: Miller & Thode (2007)."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.caption     = element_text(color = "grey50", size = 9),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(size = 11),
    axis.text        = element_text(size = 10)
  )

print(p4a)
print(p4b)
print(p4c)

# --- QA: confirm non-empty plots and reasonable value ranges -----------------
stopifnot("dNBR histogram has no data"  = nrow(cbi_west) > 0)
stopifnot("RdNBR histogram has no data" = nrow(rdnbr_df) > 0)

ggsave(here("figures/severity_dnbr_map.png"),
       plot = p4a, width = 10, height = 9, dpi = 300, bg = "white")
ggsave(here("figures/severity_dnbr_histogram.png"),
       plot = p4b, width = 9,  height = 6, dpi = 300, bg = "white")
ggsave(here("figures/severity_rdnbr_histogram.png"),
       plot = p4c, width = 9,  height = 6, dpi = 300, bg = "white")

img_dims <- dim(png::readPNG(here("figures/severity_dnbr_map.png")))
cat("Plot 4a saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
img_dims <- dim(png::readPNG(here("figures/severity_dnbr_histogram.png")))
cat("Plot 4b saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
img_dims <- dim(png::readPNG(here("figures/severity_rdnbr_histogram.png")))
cat("Plot 4c saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

## Summary

```{r summary-table}
# Quick summary table of what was found — useful for NOTES.md
n_valid_thresh <- sum(mtbs_west$thresholds_valid, na.rm = TRUE)
median_dnbr    <- median(cbi_west$dnbr_val, na.rm = TRUE)
median_rdnbr   <- median(rdnbr_df$rdnbr,    na.rm = TRUE)

summary_tbl <- tibble::tribble(
  ~Metric,                                ~Value,
  "Total wildfires (2000–2023)",          scales::comma(n_fires_total),
  "States covered",                       paste(present_states, collapse = ", "),
  "Median fire size",                     paste(scales::comma(size_stats$median_ac), "acres"),
  "Fires with valid severity thresholds", scales::comma(n_valid_thresh),
  "CBI field plots (Western US)",         scales::comma(nrow(cbi_west)),
  "Median dNBR (CBI plots)",             as.character(round(median_dnbr)),
  "Median RdNBR (CBI plots)",            as.character(round(median_rdnbr))
)

knitr::kable(summary_tbl, col.names = c("Metric", "Value"), align = "lr")
```

---

## Step 3 — GEE Biomass Extraction

```{r biomass-setup}
START_YEAR_BIO <- 1984
END_YEAR_BIO   <- 2023
STATE_LABEL    <- "California"
CSV_PATH       <- here("output/biomass_timeseries.csv")
PYTHON_SCRIPT  <- here("scripts/python/01_extract_biomass_gee.py")

dir.create(here("output"), showWarnings = FALSE)
```

Run the Python extraction script if the CSV has not been generated yet. Requires prior `earthengine authenticate` in a terminal.

```{r run-extraction}
if (!file.exists(CSV_PATH)) {
  message("Biomass CSV not found — running GEE extraction for ", STATE_LABEL, "...")
  python_paths <- c(
    "C:/Users/shaht/anaconda3/python.exe",
    Sys.which("python3"),
    Sys.which("python")
  )
  python_exe <- python_paths[file.exists(python_paths)][1]
  if (is.na(python_exe)) stop("Python not found. Set the path manually in this chunk.")
  result <- system2(python_exe, args = PYTHON_SCRIPT, stdout = TRUE, stderr = TRUE)
  cat(result, sep = "\n")
  if (!file.exists(CSV_PATH)) {
    stop("Extraction did not produce output/biomass_timeseries.csv.\n",
         "Check GEE authentication and GEE_PROJECT in 01_extract_biomass_gee.py.")
  }
  message("Extraction complete.")
} else {
  message("CSV found at ", CSV_PATH, " — skipping extraction.")
}
```

```{r load-biomass}
biomass_raw <- read_csv(CSV_PATH, show_col_types = FALSE)

stopifnot(
  "Missing expected columns" =
    all(c("event_id", "fire_year", "year", "fitted_nbr") %in% names(biomass_raw))
)
stopifnot("Year range out of bounds" =
    all(biomass_raw$year >= START_YEAR_BIO & biomass_raw$year <= END_YEAR_BIO))

n_fires_bio <- n_distinct(biomass_raw$event_id)
pct_valid   <- mean(!is.na(biomass_raw$fitted_nbr)) * 100

cat("Fires:      ", n_fires_bio, "\n")
cat("Year range: ", min(biomass_raw$year), "–", max(biomass_raw$year), "\n")
cat("Valid NBR:  ", round(pct_valid, 1), "%\n")

if (pct_valid < 50) {
  warning("< 50% valid NBR — check date window and Landsat coverage.")
}
```

```{r prepare-biomass}
# Convert from LandTrendR inverted space (loss = positive) to display scale
# (higher = more vegetation = more biomass)
biomass <- biomass_raw |>
  mutate(display_nbr = -fitted_nbr) |>
  filter(!is.na(display_nbr))

cat("Records retained after NA removal:", nrow(biomass), "\n")

# Single-fire sanity check: trajectory should show a dip at the fire year
example_fire <- biomass |>
  filter(event_id == first(unique(event_id))) |>
  arrange(year)

ggplot(example_fire, aes(x = year, y = display_nbr)) +
  geom_line(color = "#2c7bb6") +
  geom_point(color = "#2c7bb6") +
  geom_vline(xintercept = first(example_fire$fire_year),
             linetype = "dashed", color = "#d73027") +
  annotate("text", x = first(example_fire$fire_year) + 0.3, y = Inf,
           vjust = 1.5, hjust = 0, label = "Fire year",
           color = "#d73027", size = 3.5) +
  labs(title = glue("Sanity check: NBR trajectory for {first(example_fire$event_id)}"),
       x = "Year", y = "NBR (display scale — higher = more vegetation)") +
  theme_minimal(base_size = 11)
```

---

## Plot 5 — Annual Mean NBR Over Time

```{r plot5-biomass-trend}
annual_nbr <- biomass |>
  group_by(year) |>
  summarise(
    mean_nbr = mean(display_nbr),
    se_nbr   = sd(display_nbr) / sqrt(n()),
    n_fires  = n_distinct(event_id),
    .groups  = "drop"
  )

stopifnot("Annual NBR should not be all NA" = !all(is.na(annual_nbr$mean_nbr)))

p5 <- ggplot(annual_nbr, aes(x = year, y = mean_nbr)) +
  geom_ribbon(aes(ymin = mean_nbr - se_nbr, ymax = mean_nbr + se_nbr),
              fill = "#2c7bb6", alpha = 0.2) +
  geom_line(color = "#2c7bb6", linewidth = 1) +
  geom_point(color = "#2c7bb6", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE,
              color = "black", linetype = "dashed", linewidth = 0.8) +
  scale_x_continuous(
    breaks = seq(START_YEAR_BIO, END_YEAR_BIO, by = 5),
    expand = c(0.01, 0)
  ) +
  scale_y_continuous(labels = number_format(accuracy = 0.001)) +
  labs(
    title    = glue("Annual Mean NBR — {STATE_LABEL} Wildfire Sites ({START_YEAR_BIO}–{END_YEAR_BIO})"),
    subtitle = glue("Raw annual Landsat composites across {n_fires_bio} sampled MTBS fire perimeters"),
    x        = "Year",
    y        = "Mean NBR (display scale — higher = more vegetation)",
    caption  = paste(
      "Shaded band = ± 1 SE. Dashed line = linear trend.",
      "Source: USGS MTBS + Landsat via Google Earth Engine."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey40", size = 11),
    plot.caption     = element_text(color = "grey50", size = 9),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(size = 11),
    axis.text        = element_text(size = 10)
  )

print(p5)

stopifnot("Y-axis is all zero — check NBR sign" = !all(annual_nbr$mean_nbr == 0))

ggsave(here("figures/biomass_trend_over_time.png"),
       plot = p5, width = 10, height = 6, dpi = 300, bg = "white")
img_dims <- dim(png::readPNG(here("figures/biomass_trend_over_time.png")))
cat("Plot 5 saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

## Plot 6 — Biomass Before and After Fire Events

```{r plot6-before-after}
biomass_event <- biomass |>
  mutate(rel_year = year - fire_year) |>
  filter(rel_year >= -5, rel_year <= 10)

event_mean <- biomass_event |>
  group_by(rel_year) |>
  summarise(
    mean_nbr = mean(display_nbr),
    se_nbr   = sd(display_nbr) / sqrt(n()),
    n_fires  = n_distinct(event_id),
    .groups  = "drop"
  )

cat("Fires contributing at rel_year =  0:",
    event_mean$n_fires[event_mean$rel_year == 0], "\n")
cat("Fires contributing at rel_year = -5:",
    event_mean$n_fires[event_mean$rel_year == -5], "\n")
cat("Fires contributing at rel_year = +10:",
    event_mean$n_fires[event_mean$rel_year == 10], "\n")

p6 <- ggplot(event_mean, aes(x = rel_year, y = mean_nbr)) +
  geom_ribbon(aes(ymin = mean_nbr - se_nbr, ymax = mean_nbr + se_nbr),
              fill = "#d73027", alpha = 0.2) +
  geom_line(color = "#d73027", linewidth = 1) +
  geom_point(color = "#d73027", size = 2.5) +
  geom_vline(xintercept = 0,
             linetype = "dashed", linewidth = 0.9, color = "grey30") +
  annotate("text", x = 0.2, y = Inf, vjust = 1.5, hjust = 0,
           label = "Fire year", color = "grey30", size = 3.5) +
  scale_x_continuous(
    breaks = -5:10,
    labels = function(x) ifelse(x == 0, "0\n(fire)", as.character(x))
  ) +
  scale_y_continuous(labels = number_format(accuracy = 0.001)) +
  labs(
    title    = glue("Biomass Before and After Wildfire Events — {STATE_LABEL}"),
    subtitle = "Event-study window: −5 to +10 years relative to fire ignition",
    x        = "Years relative to fire",
    y        = "Mean NBR (display scale)",
    caption  = paste(
      "Each fire contributes one trajectory centered on its ignition year.",
      "Shaded band = ± 1 SE.",
      "Source: USGS MTBS + Landsat via Google Earth Engine."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey40", size = 11),
    plot.caption     = element_text(color = "grey50", size = 9),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(size = 11),
    axis.text        = element_text(size = 10)
  )

print(p6)

stopifnot("Year 0 must have data" = 0 %in% event_mean$rel_year)
stopifnot("CI should not be wider than ± 0.5 NBR" =
    all(event_mean$se_nbr < 0.5, na.rm = TRUE))

ggsave(here("figures/biomass_before_after_fire.png"),
       plot = p6, width = 10, height = 6, dpi = 300, bg = "white")
img_dims <- dim(png::readPNG(here("figures/biomass_before_after_fire.png")))
cat("Plot 6 saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

## Combined Biomass View

```{r combined-plot, fig.height=12}
combined <- (p5 + labs(subtitle = NULL, caption = NULL)) /
            (p6 + labs(subtitle = NULL, caption = NULL)) +
  plot_annotation(
    title   = glue("Wildfire Biomass EDA — {STATE_LABEL}"),
    caption = "Source: USGS MTBS + Landsat via Google Earth Engine. Raw annual composites."
  )

print(combined)

ggsave(here("figures/biomass_combined.png"),
       plot = combined, width = 10, height = 12, dpi = 300, bg = "white")
```
