---
title: "Biomass Within Fire Perimeters — EDA"
subtitle: "ctrees vs. eMapR within MTBS fire polygons | California wildfire cohorts"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 6
    theme: flatly
params:
  study_year_min: 2000   # first fire cohort year
  study_year_max: 2005   # last fire cohort year
execute:
  warning: false
  message: false
  cache: false
---

This document explores aboveground biomass (AGB) **within MTBS fire perimeters** for
California fires that ignited in `r params$study_year_min`–`r params$study_year_max`. It compares
two biomass datasets — **ctrees** (annual, ML-based) and **eMapR** (annual,
LandTrendR-based) — using per-fire mean AGB extracted directly from raster and tabular
data. No Google Earth Engine extraction is required; all data are local.

```{r outline}
# =============================================================================
# SCRIPT OUTLINE
# 1.  Setup — packages, study parameters (from params$study_year_min/max), paths, theme
# 2.  Study fire cohort — MTBS perimeters for STUDY_YEARS CA fires; map + table
# 3.  Biomass within fires — ctrees (load CSV, distributions, trend by year)
# 4.  Biomass within fires — eMapR (extract from ~100 m TIFs, cache to CSV)
# 5.  Dataset comparison — distributions, scatter, summary table
# 6.  Before / after fire event study (EDA_PLAN §4.2)
# 7.  Design notes — data limitations and implications
# =============================================================================
```

---

## 1. Setup

```{r setup}
library(terra)
library(sf)
library(here)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)
library(glue)
library(patchwork)
library(tigris)
library(lubridate)
library(ggspatial)
library(knitr)
library(stringr)
library(RColorBrewer)
library(viridis)
library(plotly)

sf_use_s2(FALSE)
options(tigris_use_cache = TRUE, knitr.kable.NA = "—")
here::i_am("analysis/05_biomass_within_fires.qmd")

# ── Study parameters ──────────────────────────────────────────────────────────
STUDY_YEARS <- as.integer(seq(params$study_year_min, params$study_year_max))
STATE_FIPS  <- "CA"

cat("Study cohort years:", params$study_year_min, "–", params$study_year_max,
    "(", length(STUDY_YEARS), "cohorts )\n")

# ── Paths ─────────────────────────────────────────────────────────────────────
MTBS_PATH       <- here("data", "raw", "mtbs", "mtbs_perimeter_data", "mtbs_perims_DD.shp")
EMAPR_DIR       <- here("data", "processed", "emapr_biomass_ca")
CTREES_FIRE_CSV <- here("data", "processed", "ctrees",
                        "biomass_fire_polygons_ctrees_forested.csv")

# Cache keyed by cohort year range — "2000_2015" instead of listing all 16 years.
years_hash     <- paste(params$study_year_min, params$study_year_max, sep = "_")
EMAPR_FIRE_CSV <- here("data", "processed", "emapr_biomass_ca",
                       glue("biomass_fire_polygons_emapr_{years_hash}_100m_forested.csv"))

dir.create(here("figures"), showWarnings = FALSE)

# Detect all 30 m CA-clipped source TIFs (1990–2015 currently on disk).
# These determine which years can be extracted.
# ~100 m versions (~55 MB each, fact=3 from 30 m) are used for extraction —
# run scripts/r/01_create_emapr_100m_tifs.R once to generate them.
emapr_src_tifs <- list.files(EMAPR_DIR, pattern = "^composite_\\d{4}_ca\\.tif$",
                             full.names = FALSE)
EMAPR_YEARS    <- sort(as.integer(str_extract(emapr_src_tifs, "\\d{4}")))
cat("eMapR CA source TIFs (30 m) available for:", paste(EMAPR_YEARS, collapse = ", "), "\n")

# ── CA boundary ───────────────────────────────────────────────────────────────
ca_sf   <- tigris::states(cb = TRUE, year = 2022, resolution = "5m") |>
  dplyr::filter(STUSPS == STATE_FIPS)
stopifnot("State boundary not found" = nrow(ca_sf) == 1)

ca_5070 <- sf::st_transform(ca_sf, 5070)

# ── Shared ggplot theme ───────────────────────────────────────────────────────
theme_baci <- theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey40", size = 11),
    plot.caption     = element_text(color = "grey50", size = 9),
    panel.grid.minor = element_blank(),
    axis.text        = element_text(size = 10)
  )

# ── Cohort colors — viridis gradient, works for any number of study years ─────
COHORT_COLORS <- setNames(
  viridis::viridis(length(STUDY_YEARS), option = "D"),
  as.character(STUDY_YEARS)
)
```

---

## 2. Study Fire Cohort — MTBS Perimeters

```{r mtbs-load}
# ── Load and filter MTBS to study fires ──────────────────────────────────────
mtbs_raw <- sf::st_read(MTBS_PATH, quiet = TRUE)
cat("Total MTBS fires loaded:        ", nrow(mtbs_raw), "\n")

mtbs_wf <- mtbs_raw |>
  dplyr::filter(incid_type == "Wildfire")
cat("After Wildfire filter:          ", nrow(mtbs_wf), "\n")

mtbs_study_all <- mtbs_wf |>
  dplyr::mutate(year = lubridate::year(ig_date)) |>
  dplyr::filter(year %in% STUDY_YEARS)
cat("After study-year filter:        ", nrow(mtbs_study_all), "\n")

mtbs_study <- mtbs_study_all |>
  sf::st_transform(5070) |>
  sf::st_filter(ca_5070) |>
  # Keep only fires whose MTBS event_id starts with the study state code.
  # sf::st_filter() uses perimeter intersection, so border fires from adjacent
  # states (e.g. AZ, NV) that clip into CA are included geometrically but are
  # not CA fires — drop them here.
  dplyr::filter(startsWith(event_id, STATE_FIPS)) |>
  dplyr::filter(burnbndac >= 1000) |>
  dplyr::mutate(
    asmnt_cat = dplyr::case_when(
      grepl("Extended",  asmnt_type) ~ "Extended",
      grepl("Initial",   asmnt_type) ~ "Initial",
      grepl("Emergency", asmnt_type) ~ "Emergency",
      TRUE                           ~ "Other"
    )
  )
cat("After CA spatial filter + 1000ac:", nrow(mtbs_study), "fires\n")

# Geometry validity check
invalid <- sum(!sf::st_is_valid(mtbs_study))
if (invalid > 0) {
  warning(glue("{invalid} invalid geometries — fixing with st_make_valid()"))
  mtbs_study <- sf::st_make_valid(mtbs_study)
}

# No duplicate event_ids
stopifnot("Duplicate event_ids found" = !any(duplicated(mtbs_study$event_id)))
cat("Total burned area:               ",
    scales::comma(round(sum(mtbs_study$burnbndac))), "acres\n")
```

### 2.1 Map of Study Fires

```{r study-fires-map, fig.height=9}
# CA boundary as path for overlay
ca_path_5070 <- sf::st_coordinates(sf::st_cast(ca_5070, "MULTILINESTRING")) |>
  as.data.frame() |>
  dplyr::mutate(grp = paste(L1, L2, sep = "_"))

ggplot() +
  geom_sf(data = ca_5070, fill = "grey95", color = "grey60", linewidth = 0.4) +
  geom_sf(data = mtbs_study,
          aes(fill = factor(year)), color = NA, alpha = 0.80) +
  scale_fill_manual(
    values = COHORT_COLORS,
    name   = "Ignition year"
  ) +
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(location = "tr", which_north = "true",
                         style = north_arrow_minimal()) +
  labs(
    title    = glue("MTBS Study Fire Cohorts — {STATE_FIPS}"),
    subtitle = glue("{nrow(mtbs_study)} wildfires ≥ 1,000 acres, ignition years {min(STUDY_YEARS)}–{max(STUDY_YEARS)}"),
    caption  = "Source: USGS MTBS Program (mtbs.gov)."
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(color = "grey40", size = 11),
    plot.caption    = element_text(color = "grey50", size = 9),
    legend.position = "bottom"
  )
```

### 2.2 Study Fire Summary

```{r study-fires-summary}
# ── Pivot wide: one row per year, separate columns for Extended and Initial ────
fire_summary_wide <- mtbs_study |>
  sf::st_drop_geometry() |>
  dplyr::group_by(year, asmnt_cat) |>
  dplyr::summarise(
    n_fires   = dplyr::n(),
    median_ac = round(median(burnbndac)),
    total_ac  = round(sum(burnbndac)),
    .groups   = "drop"
  ) |>
  dplyr::filter(asmnt_cat %in% c("Extended", "Initial")) |>
  tidyr::pivot_wider(
    names_from  = asmnt_cat,
    values_from = c(n_fires, median_ac, total_ac)
  ) |>
  dplyr::arrange(year)

# ── Helper: stack Extended (red) over Initial (blue) in one HTML cell ─────────
ext_col <- "#d6604d"
ini_col <- "#4393c3"

html_cell <- function(ext_val, ini_val) {
  fmt <- function(x) ifelse(is.na(x), "&mdash;", scales::comma(x))
  glue(
    '<span style="color:{ext_col}">{fmt(ext_val)}</span><br>',
    '<span style="color:{ini_col}">{fmt(ini_val)}</span>'
  )
}

# ── Collapse to 4-column table ────────────────────────────────────────────────
fire_summary_html <- fire_summary_wide |>
  dplyr::mutate(
    year      = as.character(year),
    n_fires   = html_cell(n_fires_Extended,   n_fires_Initial),
    median_ac = html_cell(median_ac_Extended, median_ac_Initial),
    total_ac  = html_cell(total_ac_Extended,  total_ac_Initial)
  ) |>
  dplyr::select(year, n_fires, median_ac, total_ac)

knitr::kable(
  fire_summary_html,
  col.names = c("Year", "N fires", "Median (ac)", "Total (ac)"),
  escape    = FALSE,
  align     = c("l", "r", "r", "r"),
  caption   = paste0(
    "Study fire cohort summary by ignition year. ",
    '<span style="color:', ext_col, '">&#9632; Extended</span>',
    " &nbsp; ",
    '<span style="color:', ini_col, '">&#9632; Initial</span>.'
  )
)
```

---

## 3. Biomass Within Fires — ctrees

ctrees AGB has already been extracted within MTBS fire perimeters and stored in
`biomass_fire_polygons_ctrees.csv` (columns: `event_id`, `fire_year`, `year`,
`agb_mean_mgha`). Each row is the mean AGB within one fire polygon for one
observation year.

```{r ctrees-load}
# ── Load and filter to study fires ────────────────────────────────────────────
# Fall back to the non-forested CSV while the slow forested extraction is running.
# Both have identical columns (event_id, fire_year, year, agb_mean_mgha);
# the only difference is the forested version masks to NLCD 2004 forest pixels.
CTREES_NONFORESTED_CSV <- here("data", "processed", "ctrees",
                               "biomass_fire_polygons_ctrees.csv")

if (file.exists(CTREES_FIRE_CSV)) {
  ctrees_src <- CTREES_FIRE_CSV
  cat("Using forested ctrees CSV (forest-masked).\n")
} else if (file.exists(CTREES_NONFORESTED_CSV)) {
  ctrees_src <- CTREES_NONFORESTED_CSV
  message("Forested CSV not found — falling back to non-forested ctrees CSV.\n",
          "Run scripts/r/04_extract_ctrees_within_fires.R then re-render.")
} else {
  stop("No ctrees CSV found. Run scripts/r/04_extract_ctrees_within_fires.R first.")
}

ctrees_all   <- readr::read_csv(ctrees_src, show_col_types = FALSE)
ctrees_fires <- ctrees_all |>
  dplyr::filter(event_id %in% mtbs_study$event_id) |>
  dplyr::mutate(
    rel_year  = year - fire_year,
    fire_year = as.integer(fire_year)
  ) |>
  dplyr::filter(!is.na(agb_mean_mgha))

cat("Source CSV:                ", basename(ctrees_src), "\n")
cat("Rows in full ctrees CSV:   ", nrow(ctrees_all), "\n")
cat("Rows for study fires:      ", nrow(ctrees_fires), "\n")
cat("Fires matched to MTBS:     ",
    dplyr::n_distinct(ctrees_fires$event_id), "of", nrow(mtbs_study), "\n")
cat("Observation years:         ",
    min(ctrees_fires$year), "–", max(ctrees_fires$year), "\n")
cat("rel_year range:            ",
    min(ctrees_fires$rel_year), "to", max(ctrees_fires$rel_year), "\n")

stopifnot("fire_year values outside STUDY_YEARS" =
            all(ctrees_fires$fire_year %in% STUDY_YEARS))
```

---

## 4. Biomass Within Fires — eMapR

eMapR AGB is extracted here from the 300 m CA-clipped TIFs using `terra::extract()`.
Results are cached to a CSV keyed by study year range — switching `params$study_year_min`/`max`
automatically uses (or creates) the correct cache file.

```{r emapr-inventory}
# ── Inventory available ~100 m TIFs ───────────────────────────────────────────
# ~100 m TIFs (~55 MB each, fact=3 from 30 m) must be pre-generated before
# rendering. Run scripts/r/01_create_emapr_100m_tifs.R once to create them
# for all years. These files are memory-safe for terra::extract() and match
# the ctrees native ~100 m resolution for a fair comparison.
tifs_100m_ready <- sort(as.integer(str_extract(
  list.files(EMAPR_DIR, pattern = "^composite_\\d{4}_ca_100m\\.tif$"),
  "\\d{4}"
)))
cat("~100 m TIFs available for:", paste(tifs_100m_ready, collapse = ", "), "\n")

missing_100m <- setdiff(EMAPR_YEARS, tifs_100m_ready)
if (length(missing_100m) > 0) {
  message(
    length(missing_100m), " year(s) have no ~100 m TIF yet and will be skipped: ",
    paste(missing_100m, collapse = ", "),
    "\nRun scripts/r/01_create_emapr_100m_tifs.R to generate them."
  )
}
```

```{r emapr-extract}
# ── Load from pre-built cache only ────────────────────────────────────────────
# Extraction runs in scripts/r/02_extract_emapr_within_fires.R (run once in
# the R console before rendering). Doing terra::extract() inside Quarto causes
# invisible hangs on Windows because Quarto buffers chunk output and terra's
# internal C++ threading can deadlock.
EMAPR_AVAILABLE <- file.exists(EMAPR_FIRE_CSV)

if (!EMAPR_AVAILABLE) {
  message("eMapR cache not found (", basename(EMAPR_FIRE_CSV), ").\n",
          "Run scripts/r/02_extract_emapr_within_fires.R to enable eMapR sections.\n",
          "Continuing with ctrees only.")
  emapr_fires <- dplyr::tibble(event_id = character(), fire_year = integer(),
                               year = integer(), agb_mean_mgha = double(),
                               rel_year = integer())
} else {
  emapr_raw    <- readr::read_csv(EMAPR_FIRE_CSV, show_col_types = FALSE)
  cached_years <- sort(unique(emapr_raw$year))
  stale_years  <- setdiff(tifs_100m_ready, cached_years)

  if (length(stale_years) > 0) {
    message(length(stale_years), " year(s) available in TIFs but not cached: ",
            paste(stale_years, collapse = ", "),
            "\nDelete cache and re-run 02_extract_emapr_within_fires.R to refresh.")
  }

  cat("Cache loaded —", length(cached_years), "years,", nrow(emapr_raw), "rows.\n")
  cat("Cache file:", basename(EMAPR_FIRE_CSV), "\n")

  fire_meta <- sf::st_drop_geometry(mtbs_study) |>
    dplyr::select(event_id, fire_year = year)

  emapr_fires <- emapr_raw |>
    dplyr::left_join(fire_meta, by = "event_id") |>
    dplyr::mutate(
      rel_year  = year - fire_year,
      fire_year = as.integer(fire_year)
    ) |>
    dplyr::filter(!is.na(agb_mean_mgha))

  cat("eMapR rows for study fires:    ", nrow(emapr_fires), "\n")
  cat("Fires with eMapR data:         ",
      dplyr::n_distinct(emapr_fires$event_id), "of", nrow(mtbs_study), "\n")
  cat("Observation years:             ",
      min(emapr_fires$year), "–", max(emapr_fires$year), "\n")
  cat("rel_year range:                ",
      min(emapr_fires$rel_year), "to", max(emapr_fires$rel_year), "\n")
}
```

### 4.1 Mean Annual AGB — ctrees vs. eMapR

```{r combined-agb-by-year, fig.height=5}
# ── Pool mean AGB across all study fires per year, one line per dataset ───────
# Both ctrees_fires and emapr_fires are defined; this chunk runs after emapr-extract
combined_yr <- dplyr::bind_rows(
  ctrees_fires |>
    dplyr::group_by(year) |>
    dplyr::summarise(mean_agb = mean(agb_mean_mgha, na.rm = TRUE), .groups = "drop") |>
    dplyr::mutate(dataset = "ctrees (~100 m)"),
  emapr_fires |>
    dplyr::group_by(year) |>
    dplyr::summarise(mean_agb = mean(agb_mean_mgha, na.rm = TRUE), .groups = "drop") |>
    dplyr::mutate(dataset = "eMapR (~100 m)")
)

ggplot(combined_yr, aes(x = year, y = mean_agb, color = dataset)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_vline(xintercept = STUDY_YEARS, linetype = "dashed",
             color = "grey50", linewidth = 0.5) +
  annotate("text", x = STUDY_YEARS + 0.2, y = Inf,
           label = paste0("Fire\n", STUDY_YEARS),
           vjust = 1.4, hjust = 0, size = 3, color = "grey40") +
  scale_color_manual(
    values = c("ctrees (~100 m)" = "#2166ac", "eMapR (~100 m)" = "#d6604d"),
    name   = "Dataset"
  ) +
  scale_x_continuous(breaks = seq(min(combined_yr$year), max(combined_yr$year), by = 1)) +
  scale_y_continuous(labels = scales::comma_format(),
                     expand = expansion(mult = c(0.05, 0.15))) +
  labs(
    title    = "Mean Annual AGB Within Study Fires — ctrees vs. eMapR",
    subtitle = glue("Mean across {nrow(mtbs_study)} CA fires igniting {min(STUDY_YEARS)}–{max(STUDY_YEARS)}"),
    x        = "Observation year",
    y        = "Mean AGB (Mg/ha)",
    caption  = "Dashed lines = fire ignition years. Source: ctrees CSV and eMapR terra::extract."
  ) +
  theme_baci
```

---

## 5. Dataset Comparison — ctrees vs. eMapR

Compare the two datasets **at the fire ignition years** (`STUDY_YEARS`). This shows
whether ctrees and eMapR agree on AGB at the moment of fire — the most directly
relevant check for the DiD study. Both datasets must have data for a given year for
it to be included.

```{r comparison-overlap-years}
# Restrict to STUDY_YEARS (the fire ignition years) that are present in both datasets.
# Using all overlapping observation years would be decoupled from the study design
# and produce an unwieldy number of facets/rows. Filtering to STUDY_YEARS keeps
# the comparison tied to the top-level parameter.
both_years <- intersect(
  STUDY_YEARS,
  intersect(unique(ctrees_fires$year), unique(emapr_fires$year))
)
cat("Study years present in both datasets:", paste(sort(both_years), collapse = ", "), "\n")
if (length(both_years) < length(STUDY_YEARS)) {
  missing <- setdiff(STUDY_YEARS, both_years)
  cat("  NOTE: no data in one or both datasets for year(s):",
      paste(missing, collapse = ", "), "\n")
}
```

### 5.1 Summary Statistics Table

```{r comparison-stats-table}
if (length(both_years) == 0) {
  message(
    "No study years have data in both datasets yet.\n",
    "Run scripts/r/01_create_emapr_100m_tifs.R to generate ~100 m TIFs for ",
    paste(setdiff(STUDY_YEARS, tifs_100m_ready), collapse = ", "),
    ", then re-render."
  )
} else {
  make_stats <- function(df, dataset_label) {
    df |>
      dplyr::filter(year %in% both_years) |>
      dplyr::group_by(year) |>
      dplyr::summarise(
        dataset  = dataset_label,
        n_fires  = dplyr::n_distinct(event_id),
        mean_agb = round(mean(agb_mean_mgha, na.rm = TRUE), 1),
        med_agb  = round(median(agb_mean_mgha, na.rm = TRUE), 1),
        sd_agb   = round(sd(agb_mean_mgha, na.rm = TRUE), 1),
        min_agb  = round(min(agb_mean_mgha, na.rm = TRUE), 1),
        max_agb  = round(max(agb_mean_mgha, na.rm = TRUE), 1),
        .groups  = "drop"
      )
  }

  stats_df <- dplyr::bind_rows(
    make_stats(ctrees_fires, "ctrees"),
    make_stats(emapr_fires,  "eMapR")
  ) |>
    dplyr::arrange(year, dataset)

  print(knitr::kable(
    stats_df,
    col.names   = c("Year", "Dataset", "N fires", "Mean (Mg/ha)", "Median",
                    "SD", "Min", "Max"),
    format.args = list(big.mark = ","),
    caption     = glue("AGB summary statistics within study fire polygons at each ",
                       "fire ignition year ({min(both_years)}–{max(both_years)}).")
  ))
}
```

### 5.2 Distribution Comparison

```{r comparison-density, fig.height=5}
if (length(both_years) == 0) {
  message("Skipping distribution plot — no overlapping study years available.")
} else {
  combined_dist <- dplyr::bind_rows(
    ctrees_fires |>
      dplyr::filter(year %in% both_years) |>
      dplyr::mutate(dataset = "ctrees"),
    emapr_fires |>
      dplyr::filter(year %in% both_years) |>
      dplyr::mutate(dataset = "eMapR")
  )

  # Boxplot by cohort year — readable for 16 years; density facets would overflow
  print(
    ggplot(combined_dist,
           aes(x = factor(year), y = agb_mean_mgha,
               fill = dataset, color = dataset)) +
      geom_boxplot(alpha = 0.35, linewidth = 0.5, outlier.size = 0.8,
                   position = position_dodge(width = 0.75)) +
      scale_fill_manual(values  = c("ctrees" = "#2166ac", "eMapR" = "#d6604d"),
                        name    = "Dataset") +
      scale_color_manual(values = c("ctrees" = "#2166ac", "eMapR" = "#d6604d"),
                         name   = "Dataset") +
      scale_y_continuous(labels = scales::comma_format()) +
      labs(
        title    = "AGB Distribution Within Study Fires — ctrees vs. eMapR",
        subtitle = glue("Per-fire mean AGB at each cohort's ignition year ({min(both_years)}–{max(both_years)})"),
        x        = "Fire cohort year",
        y        = "Mean AGB within fire perimeter (Mg/ha)",
        caption  = "Each box = distribution across fires in that cohort year (~100 m pixels for both datasets)."
      ) +
      theme_baci +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

### 5.3 Scatter: ctrees vs. eMapR (per fire × year)

```{r comparison-scatter, fig.height=6}
if (length(both_years) == 0) {
  message("Skipping scatter plot — no overlapping study years available.")
} else {
  scatter_df <- dplyr::inner_join(
    ctrees_fires |>
      dplyr::filter(year %in% both_years) |>
      dplyr::select(event_id, year, fire_year, ctrees_agb = agb_mean_mgha),
    emapr_fires |>
      dplyr::filter(year %in% both_years) |>
      dplyr::select(event_id, year, emapr_agb = agb_mean_mgha),
    by = c("event_id", "year")
  )

  r_val  <- round(cor(scatter_df$ctrees_agb, scatter_df$emapr_agb,
                      use = "complete.obs"), 3)
  rmse   <- round(sqrt(mean((scatter_df$ctrees_agb - scatter_df$emapr_agb)^2,
                            na.rm = TRUE)), 1)
  n_pts  <- nrow(scatter_df)
  xy_max <- max(scatter_df$ctrees_agb, scatter_df$emapr_agb, na.rm = TRUE) * 1.05
  cat(glue("Pearson r = {r_val} | RMSE = {rmse} Mg/ha | N = {n_pts} fire–year pairs\n"))

  print(
    ggplot(scatter_df, aes(x = ctrees_agb, y = emapr_agb,
                           color = factor(fire_year))) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                  color = "grey50", linewidth = 0.8) +
      geom_point(alpha = 0.65, size = 2) +
      geom_smooth(method = "lm", se = FALSE, linewidth = 0.9,
                  aes(group = 1), color = "grey20") +
      annotate("text", x = 0.05 * xy_max, y = 0.92 * xy_max,
               label = glue("r = {r_val}\nRMSE = {rmse} Mg/ha\nn = {n_pts}"),
               hjust = 0, size = 3.5, color = "grey25") +
      scale_color_manual(values = COHORT_COLORS, name = "Fire cohort") +
      scale_x_continuous(limits = c(0, xy_max), labels = scales::comma_format()) +
      scale_y_continuous(limits = c(0, xy_max), labels = scales::comma_format()) +
      coord_fixed() +
      labs(
        title    = "ctrees vs. eMapR — Per-fire Mean AGB",
        subtitle = "One point = one fire × one overlapping year | dashed = 1:1 line",
        x        = "ctrees mean AGB (Mg/ha)",
        y        = "eMapR mean AGB (Mg/ha)",
        caption  = "Source: ctrees CSV and eMapR terra::extract."
      ) +
      theme_baci
  )
}
```

---

## 6. Before / After Fire Event Study

This section adapts EDA_PLAN §4.2 to use actual AGB (Mg/ha) from both datasets,
indexed to each fire's ignition year (`rel_year = 0`).

```{r event-study-data}
# ── Build combined long data frame indexed to fire year ───────────────────────
REL_MIN <- -12L   # years before fire to show (eMapR reaches back to 1990 for 2002 cohort)
REL_MAX <-  15L   # years after fire to show (2000 cohort + 15 = 2015, last eMapR year)

event_long <- dplyr::bind_rows(
  ctrees_fires |> dplyr::mutate(dataset = "ctrees (~100 m)"),
  emapr_fires  |> dplyr::mutate(dataset = "eMapR (~100 m)")
) |>
  dplyr::filter(rel_year >= REL_MIN, rel_year <= REL_MAX)

# Aggregate: mean ± SE across fires, by dataset × rel_year
event_agg <- event_long |>
  dplyr::group_by(dataset, rel_year) |>
  dplyr::summarise(
    mean_agb = mean(agb_mean_mgha, na.rm = TRUE),
    se_agb   = sd(agb_mean_mgha,   na.rm = TRUE) / sqrt(dplyr::n()),
    n_fires  = dplyr::n_distinct(event_id),
    .groups  = "drop"
  )

cat("Event-study rows per dataset:\n")
print(dplyr::count(event_agg, dataset))
```

### 6.1 Pooled Event Study — All Cohorts Combined

```{r event-study-plot, fig.height=5}
n_fires_total  <- dplyr::n_distinct(event_long$event_id)

# Show the full window where data exists.
# Pre-fire: eMapR reaches back to 1990 (rel_year = -12 for 2002 cohort);
#   ctrees starts at 2000 so has no pre-fire coverage for 2000-cohort.
# Post-fire: eMapR runs to 2015 (rel_year = +15 for 2000 cohort);
#   ctrees continues to 2025 but we cap at +15 to keep axes matched.
# Pooling cohorts at extreme rel_years means fewer fires contribute — note this
# in interpretation (2002-cohort dominates rel_year < -10; 2000-cohort dominates > +13).
event_agg_pooled <- dplyr::filter(event_agg, rel_year >= REL_MIN, rel_year <= REL_MAX)

# Build the tooltip text on the point layer; other layers get no tooltip
p_event <- ggplot(event_agg_pooled, aes(x = rel_year, y = mean_agb,
                                 color = dataset)) +
  geom_line(linewidth = 1.0) +
  geom_point(
    aes(text = glue(
      "Dataset: {dataset}\n",
      "Rel. year: {rel_year}\n",
      "Mean AGB: {round(mean_agb, 1)} Mg/ha\n",
      "±1 SE: ±{round(se_agb, 1)} Mg/ha\n",
      "N fires: {n_fires}"
    )),
    size = 2.5
  ) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey30", linewidth = 0.8) +
  scale_color_manual(
    values = c("ctrees (~100 m)" = "#2166ac", "eMapR (~100 m)" = "#d6604d"),
    name   = "Dataset"
  ) +
  scale_x_continuous(
    breaks = seq(REL_MIN, REL_MAX, by = 2),
    name   = "Years relative to fire"
  ) +
  scale_y_continuous(
    labels = scales::comma_format(),
    expand = expansion(mult = c(0.05, 0.15)),
    name   = "Mean AGB (Mg/ha)"
  ) +
  labs(
    title  = "Aboveground Biomass Before and After Fire — ctrees vs. eMapR",
    x      = "Years relative to fire",
    y      = "Mean AGB (Mg/ha)"
  ) +
  theme_baci

plotly::ggplotly(p_event, tooltip = "text") |>
  plotly::layout(
    title = list(
      text = paste0(
        "Aboveground Biomass Before and After Fire — ctrees vs. eMapR",
        "<br><sup>Mean ±1 SE across ", n_fires_total, " CA fires igniting ",
        min(STUDY_YEARS), "–", max(STUDY_YEARS),
        " | eMapR: ", min(EMAPR_YEARS), "–2015 | ctrees: 2000–2025",
        " | dashed = fire year</sup>"
      )
    )
  )
```

### 6.2 Event Study by Fire Cohort

```{r event-study-by-cohort, fig.height=7}
# Per-cohort aggregation — use full window; each cohort's data range differs by dataset
event_agg_cohort <- event_long |>
  dplyr::group_by(dataset, fire_year, rel_year) |>
  dplyr::summarise(
    mean_agb = mean(agb_mean_mgha, na.rm = TRUE),
    se_agb   = sd(agb_mean_mgha,   na.rm = TRUE) / sqrt(dplyr::n()),
    n_fires  = dplyr::n_distinct(event_id),
    .groups  = "drop"
  )

ggplot(event_agg_cohort,
       aes(x = rel_year, y = mean_agb, color = dataset, fill = dataset)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey30", linewidth = 0.7) +
  facet_wrap(~ fire_year, ncol = 1,
             labeller = labeller(fire_year = function(x) paste("Fire cohort:", x))) +
  scale_color_manual(
    values = c("ctrees (~100 m)" = "#2166ac", "eMapR (~100 m)" = "#d6604d"),
    name   = "Dataset"
  ) +
  scale_fill_manual(
    values = c("ctrees (~100 m)" = "#2166ac", "eMapR (~100 m)" = "#d6604d"),
    name   = "Dataset"
  ) +
  scale_x_continuous(
    breaks = seq(REL_MIN, REL_MAX, by = 2),
    name   = "Years relative to fire"
  ) +
  scale_y_continuous(
    labels = scales::comma_format(),
    name   = "Mean AGB (Mg/ha)"
  ) +
  labs(
    title    = "Before / After AGB by Fire Cohort — ctrees vs. eMapR",
    subtitle = "Mean across fires in cohort | dashed = fire year (rel_year = 0) | facet = ignition year",
    caption  = "eMapR: 1990–2015 CA TIFs. ctrees: 2000–2025 CA TIFs. Source: terra::extract."
  ) +
  theme_baci +
  theme(strip.text = element_text(face = "bold"))
```

---

## 7. Design Notes

::: {.callout-note}
**ctrees pre-fire baseline**

ctrees data starts in 2000. The earliest cohort (`r min(STUDY_YEARS)`) therefore has no
ctrees pre-fire observations; the 2001 cohort has 1 year; the 2002 cohort 2 years, etc.
eMapR extends back to 1990, providing up to `r min(STUDY_YEARS) - 1990` years of
pre-fire baseline for the earliest cohort.
:::

::: {.callout-note}
**eMapR post-fire coverage**

eMapR CA TIFs currently run to 2015. The latest cohort (`r max(STUDY_YEARS)`) therefore
has 0 post-fire eMapR years; 2014 has 1 year; 2010 has 5 years; 2000 has the full
15-year recovery window. Once eMapR CONUS files for 2016–2023 are downloaded and
cropped, re-render — the cache (`biomass_fire_polygons_emapr_`r years_hash`.csv`) picks
up new TIF years automatically.
:::

::: {.callout-important}
**Implication for DiD study design**

1. Choose whether to use ctrees (better sensor stack, limited pre-period) or
   eMapR (longer pre-period, Landsat-only) as the primary outcome variable.
3. Decide whether to exclude fires that have only Initial assessments if severity
   heterogeneity is large (see §3.6 of `04_data_summary.qmd`).
:::
