---
title: "MTBS Assessment Types: Definitions and Spatial Bias"
subtitle: "Do Initial and Extended assessments cluster systematically by geography, season, or fire size?"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 6
    theme: flatly
execute:
  warning: false
  message: false
  cache: true
---

```{r}
# =============================================================================
# SCRIPT OUTLINE
# 1. Setup — packages, paths, study parameters
# 2. Assessment type definitions (comparison table)
# 3. Load and filter MTBS data
# 4. Bias investigation plots
#    4.1 Fire counts by state — Extended only + Extended/Initial stacked (patchwork)
#    4.2 Fire count by ignition month — Initial vs Extended (stacked bar)
# =============================================================================
```

---

## 1. Setup

```{r setup}
.install_if_missing <- function(pkgs) {
  missing <- pkgs[!pkgs %in% rownames(installed.packages())]
  if (length(missing) > 0) install.packages(missing, repos = "https://cloud.r-project.org", quiet = TRUE)
}
.install_if_missing(c("tigris", "ggspatial", "glue", "patchwork", "scales", "png"))

library(tidyverse)
library(sf)
library(tigris)
library(here)
library(scales)
library(glue)
library(patchwork)

options(tigris_use_cache = TRUE)
sf_use_s2(FALSE)

here::i_am("analysis/mtbs_assessment_comparison.qmd")

WESTERN_STATES <- c("AZ", "CA", "CO", "ID", "MT", "NV", "NM", "OR", "UT", "WA", "WY")
YEAR_MIN       <- 2000
YEAR_MAX       <- 2023
MTBS_PATH      <- here("data/raw/mtbs/mtbs_perimeter_data/mtbs_perims_DD.shp")

# Two-color palette: Initial (blue) vs Extended (red) — colorblind-safe
BINARY_COLORS <- c("Initial" = "#4393c3", "Extended" = "#d6604d")

dir.create(here("figures"), showWarnings = FALSE)
```

---

## 2. Assessment Type Definitions

MTBS assigns each fire an `asmnt_type` that records which post-fire Landsat image was used for burn severity mapping. The timing of that image affects the dNBR signal and the ecological interpretation of severity scores.

```{r comparison-table}
#| code-fold: false
defn_tbl <- tribble(
  ~`Assessment Type`, ~`Post-fire image timing`,                                                                         ~`Primary ecosystem`,
  "Initial",          "Acquired shortly after containment; captures first-order mortality and ash cover (Kolden 2012)", "Grasslands, shrublands",
  "Extended",         "Acquired the first growing season after fire; allows vegetation to stabilize (Kolden 2012)",     "Forests"
)

knitr::kable(defn_tbl, align = "lll")
```

## 3. Load and Filter MTBS Data

```{r load-mtbs}
mtbs_raw <- st_read(MTBS_PATH, quiet = TRUE)

stopifnot(all(c("event_id", "ig_date", "incid_type", "burnbndac",
                "asmnt_type", "pre_id", "post_id") %in% names(mtbs_raw)))

n_invalid <- sum(!st_is_valid(mtbs_raw))
if (n_invalid > 0) mtbs_raw <- st_make_valid(mtbs_raw)

cat("Records loaded:", nrow(mtbs_raw), "\n")
```

```{r parse-and-filter}
# Get Western US state boundaries for spatial filter
western_states_sf <- states(cb = TRUE, progress_bar = FALSE) |>
  filter(STUSPS %in% WESTERN_STATES) |>
  st_transform(st_crs(mtbs_raw))

# --- Parse fields -------------------------------------------------------
mtbs_parsed <- mtbs_raw |>
  mutate(
    year  = as.integer(substr(ig_date, 1, 4)),
    month = as.integer(substr(ig_date, 6, 7)),
    # State extracted from event_id prefix (format: ST + lat + lon + date)
    state = substr(event_id, 1, 2),
    # Collapse 4 types to binary: any Extended → "Extended", else → "Initial"
    # SS variants (no pre-fire image) flagged separately for the lag plot
    asmnt_binary = case_when(
      asmnt_type %in% c("Extended", "Extended (SS)") ~ "Extended",
      TRUE                                            ~ "Initial"
    ),
    asmnt_ss = str_detect(asmnt_type, fixed("(SS)"))
  )

# Filter: wildfires only, Western US, 2000–2023
n0 <- nrow(mtbs_parsed)
mtbs_wf <- mtbs_parsed |> filter(incid_type == "Wildfire")
cat("After Wildfire filter:         ", nrow(mtbs_wf), "(removed", n0 - nrow(mtbs_wf), ")\n")

mtbs_wf <- mtbs_wf |> filter(year >= YEAR_MIN, year <= YEAR_MAX)
cat("After year filter (2000–2023): ", nrow(mtbs_wf), "\n")

mtbs_study <- st_join(mtbs_wf, western_states_sf["STUSPS"], join = st_intersects, left = FALSE)
cat("After Western US spatial filter:", nrow(mtbs_study), "\n")

stopifnot(length(setdiff(WESTERN_STATES, unique(mtbs_study$STUSPS))) == 0)
cat("All 11 Western states present. ✓\n")

# Drop geometry for non-spatial summaries
mtbs_df <- st_drop_geometry(mtbs_study)

cat("\nAssessment type breakdown:\n")
print(count(mtbs_df, asmnt_type, sort = TRUE))
```

---

## 4. Bias Investigation

### 4.1 Fire Counts by State

Two views of the same data: the left plot shows Extended fires only; the right adds Initial fires as a stacked bar. Comparing them reveals both the absolute volume of Extended fires per state and how that volume relates to each state's total fire activity.

```{r plot-state-counts, fig.height=7}
state_asmnt <- mtbs_df |>
  count(STUSPS, asmnt_binary) |>
  pivot_wider(names_from = asmnt_binary, values_from = n, values_fill = 0) |>
  mutate(
    total   = Initial + Extended,
    pct_ext = 100 * Extended / total
  )

overall_pct <- 100 * sum(state_asmnt$Extended) / sum(state_asmnt$total)
cat("Overall % Extended (Western US wildfires, 2000–2023):", round(overall_pct, 1), "%\n")
print(arrange(state_asmnt, desc(Extended)))

# Shared state order: descending by Extended count
state_order <- state_asmnt |> arrange(Extended) |> pull(STUSPS)

state_long <- mtbs_df |>
  count(STUSPS, asmnt_binary) |>
  mutate(STUSPS = factor(STUSPS, levels = state_order))

# -- Plot A: Extended only ---------------------------------------------------
p_ext_only <- state_long |>
  filter(asmnt_binary == "Extended") |>
  ggplot(aes(x = n, y = STUSPS)) +
  geom_col(fill = BINARY_COLORS["Extended"], alpha = 0.85, width = 0.7) +
  geom_text(aes(label = comma(n)), hjust = -0.15, size = 3.3, color = "grey20") +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Extended only", x = "Number of fires", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title.x       = element_text(size = 11),
    axis.text          = element_text(size = 10)
  )

# -- Plot B: Extended + Initial stacked --------------------------------------
p_stacked <- state_long |>
  mutate(asmnt_binary = factor(asmnt_binary, levels = c("Initial", "Extended"))) |>
  ggplot(aes(x = n, y = STUSPS, fill = asmnt_binary)) +
  geom_col(alpha = 0.85, width = 0.7) +
  geom_text(
    data = state_asmnt |> mutate(STUSPS = factor(STUSPS, levels = state_order)),
    aes(x = total, y = STUSPS, label = comma(total), fill = NULL),
    hjust = -0.15, size = 3.3, color = "grey20"
  ) +
  scale_fill_manual(values = BINARY_COLORS, name = NULL) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Extended + Initial", x = "Number of fires", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title.x       = element_text(size = 11),
    axis.text          = element_text(size = 10),
    legend.position    = "top",
    legend.text        = element_text(size = 10)
  )

p_state_counts <- (p_ext_only | p_stacked) +
  plot_annotation(
    title    = "Wildfire Counts by State — Extended vs Initial Assessments",
    subtitle = glue("Western US wildfires ≥ 1,000 acres, 2000–2023  (n = {comma(sum(state_asmnt$total))})"),
    caption  = "States ordered by Extended fire count. Source: USGS MTBS.",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40"),
      plot.caption  = element_text(color = "grey50", size = 9)
    )
  )

print(p_state_counts)

stopifnot("All states have same Extended count — no variation" = sd(state_asmnt$Extended) > 0)

ggsave(here("figures/asmnt_counts_by_state.png"),
       plot = p_state_counts, width = 14, height = 7, dpi = 300, bg = "white")
img_dims <- dim(png::readPNG(here("figures/asmnt_counts_by_state.png")))
cat("Plot 4.1 saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

### 4.2 Fire Count by Ignition Month

Two views mirroring 4.1: Extended only on the left, stacked Extended + Initial on the right. Late-season fires (September–October) shift toward Extended because little post-fire growing season remains before winter. August has the most fires overall but a lower Extended share than September, so Initial-assessed fires dominate the high-volume peak season.

```{r plot-month-stacked, fig.height=5}
month_count <- mtbs_df |>
  filter(!is.na(month)) |>
  count(month, asmnt_binary) |>
  mutate(mon_lbl = factor(month.abb[month], levels = month.abb))

month_totals <- month_count |>
  group_by(mon_lbl) |>
  summarise(total = sum(n), .groups = "drop")

theme_month <- theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title         = element_text(size = 11),
    axis.text          = element_text(size = 10)
  )

# -- Plot A: Extended only ---------------------------------------------------
p_month_ext <- month_count |>
  filter(asmnt_binary == "Extended") |>
  ggplot(aes(x = mon_lbl, y = n)) +
  geom_col(fill = BINARY_COLORS["Extended"], alpha = 0.85, width = 0.75) +
  geom_text(aes(label = comma(n)), vjust = -0.4, size = 3.2, color = "grey20") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Extended only", x = "Ignition month", y = "Number of fires") +
  theme_month

# -- Plot B: Extended + Initial stacked --------------------------------------
p_month_stacked <- month_count |>
  mutate(asmnt_binary = factor(asmnt_binary, levels = c("Initial", "Extended"))) |>
  ggplot(aes(x = mon_lbl, y = n, fill = asmnt_binary)) +
  geom_col(alpha = 0.85, width = 0.75) +
  geom_text(
    data = month_totals,
    aes(x = mon_lbl, y = total, label = comma(total), fill = NULL),
    vjust = -0.4, size = 3.2, color = "grey20"
  ) +
  scale_fill_manual(values = BINARY_COLORS, name = NULL) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Extended + Initial", x = "Ignition month", y = "Number of fires") +
  theme_month +
  theme(legend.position = "top", legend.text = element_text(size = 10))

p_month_counts <- (p_month_ext | p_month_stacked) +
  plot_annotation(
    title    = "Wildfire Counts by Ignition Month — Extended vs Initial Assessments",
    subtitle = glue("Western US wildfires ≥ 1,000 acres, 2000–2023  (n = {comma(sum(month_totals$total))})"),
    caption  = "Labels show fire count per bar (left) or total per month (right). Source: USGS MTBS.",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40"),
      plot.caption  = element_text(color = "grey50", size = 9)
    )
  )

print(p_month_counts)

ggsave(here("figures/asmnt_counts_by_month.png"),
       plot = p_month_counts, width = 14, height = 6, dpi = 300, bg = "white")
img_dims <- dim(png::readPNG(here("figures/asmnt_counts_by_month.png")))
cat("Plot 4.2 saved — dimensions (px):", img_dims[2], "x", img_dims[1], "\n")
```

---

## Summary

```{r summary-table}
month_initial_peak <- month_count |>
  filter(asmnt_binary == "Initial") |>
  slice_max(n, n = 1)

month_pct_ext <- month_count |>
  pivot_wider(names_from = asmnt_binary, values_from = n, values_fill = 0) |>
  mutate(pct_ext = 100 * Extended / (Initial + Extended),
         mon_lbl = factor(month.abb[month], levels = month.abb)) |>
  slice_max(pct_ext, n = 1)

summary_tbl <- tribble(
  ~Metric,                                   ~Value,
  "Total Western US wildfires (2000–2023)",   comma(sum(state_asmnt$total)),
  "Extended assessments — count (%)",         glue("{comma(sum(state_asmnt$Extended))} ({round(overall_pct, 1)}%)"),
  "Highest % Extended state",                 with(state_asmnt, glue("{STUSPS[which.max(pct_ext)]} ({round(max(pct_ext), 1)}%)")),
  "Lowest % Extended state",                  with(state_asmnt, glue("{STUSPS[which.min(pct_ext)]} ({round(min(pct_ext), 1)}%)")),
  "Peak month for % Extended",                glue("{month_pct_ext$mon_lbl} ({round(month_pct_ext$pct_ext, 1)}%)"),
  "Peak month for Initial fire count",        glue("{month_initial_peak$mon_lbl} ({comma(month_initial_peak$n)} fires)")
)

knitr::kable(summary_tbl, col.names = c("Metric", "Value"), align = "lr")
```

::: {.callout-important}
Recommendation for this study (wildfire impact on biomass): 

Include both assessment types. 

The assessment type differs in what ecological changes are visible 
at the time of MTBS mapping (Key 2006), but perimeters represent 
the same burned area. Our outcome variable (annual biomass) is 
measured consistently for all fires using the same CTrees 
methodology, independent of MTBS assessment type.

Excluding initial assessments, in favor of optimizing the fire perimeter quality, would focus the analysis on forest fires only. 
:::
